---
title: "Trends in Obesity Rates Across the United States"
author: "Braydn Weaver and Adam Zhu"
date: "2024-12-08"
output: html_document
---
# Introduction

In recent years, there has been a growing interest in understanding the interplay between physical activity, nutrition, and their impacts on health outcomes such as obesity, cardiovascular fitness, and overall well-being. This data science project aims to explore trends and insights related to these critical areas by leveraging real-world data from the Nutrition, Physical Activity, and Obesity - Behavioral Risk Factor Surveillance System (BRFSS). This dataset, collected through comprehensive surveys, provides valuable information on diet, physical activity, and weight status across different demographics and geographic locations.

The primary objective of this project is to perform an exploratory data analysis (EDA) to uncover patterns and relationships between exercise habits, dietary choices, and health metrics. By analyzing this data, we aim to answer questions such as:
	•	How do physical activity levels vary across different populations?
	•	What is the relationship between exercise frequency and obesity rates?
	•	Are there significant regional differences in physical activity and nutrition trends?

The insights derived from this analysis have the potential to inform public health policies, guide community wellness programs, and raise awareness about the importance of healthy lifestyle choices. This project not only highlights the utility of data-driven approaches in addressing societal health challenges but also provides an opportunity to deepen our understanding of how daily habits influence long-term health outcomes.

# Cleaning

The dataset underwent a comprehensive cleaning process to prepare it for analysis. First, column names were standardized by replacing non-alphanumeric characters with underscores and converting all names to lowercase. Special cases, such as renaming “Age(years)” to age and “Race/Ethnicity” to race, were handled for improved clarity and consistency.

Next, duplicate rows were removed using the distinct() function to ensure each observation was unique. Columns were then converted to their appropriate data types. Numeric fields like yearstart and yearend were converted to integers, while categorical variables such as race, gender, and class were transformed into factors for efficient analysis.

Categorical variables were further standardized to clean up inconsistencies in their values. Text-based variables like gender, race, and locationdesc were converted to lowercase, stripped of leading and trailing whitespace, and formatted consistently. This step ensured uniformity across entries, preventing errors during analysis.

Finally, the cleaned dataset was saved as a new file, “cleaned_data.csv,” providing a polished version ready for exploratory data analysis. These steps ensured that the dataset was free of duplicates, had consistent column names and data types, and addressed potential issues with categorical data inconsistencies.


```{r, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(stringr)
library(readr)

data <- read_csv("Nutrition__Physical_Activity__and_Obesity_-_Behavioral_Risk_Factor_Surveillance_System.csv")
head(data)
str(data)

# Rename columns and handle special cases
data <- data %>%
  rename_with(~ str_replace_all(., "[^[:alnum:]_]", "_")) %>%  # Replace non-alphanumeric characters with underscores
  rename_with(~ str_to_lower(.)) %>% rename(
    age = "age_years_",                # Special case for "Age(years)"
    race = "race_ethnicity"           # Special case for "Race/Ethnicity"
  )

# Filter Out Irrelevant or Duplicate Rows
# Remove Duplicate Rows
data <- data %>%
  distinct()

# Convert Columns to Appropriate Data Types
data <- data %>%
  mutate(
    yearstart = as.integer(yearstart),       # Ensure year_start is integer
    yearend = as.integer(yearend),           # Ensure year_end is integer
    race = as.factor(race),                    # Convert race to factor
    gender = as.factor(gender),                # Convert gender to factor
    locationdesc  = as.factor(locationdesc),  # Convert location_desc to factor
    data_value = as.numeric(data_value),       # Ensure data_value is numeric
    class = as.factor(class),                  # Convert class to factor
    topic = as.factor(topic),                  # Convert topic to factor
    question = as.factor(question)             # Convert question to factor
  )

# Standardize Categorical Variables
# Clean up inconsistencies in categorical values
data <- data %>%
  mutate(
    gender = gender %>% str_to_lower() %>% str_trim(),  # Clean gender values
    race = race %>% str_to_lower() %>% str_trim(),      # Clean race values
    locationdesc = locationdesc %>% str_to_lower() %>% str_trim(),  # Clean location description
    class = class %>% str_to_lower() %>% str_trim(),    # Clean class values
    topic = topic %>% str_to_lower() %>% str_trim()     # Clean topic values
  )
unique(data$gender)
unique(data$race)
unique(data$locationdesc)
unique(data$class)

# 7. Save the cleaned dataset (optional)
write_csv(data, "cleaned_data.csv")
```

# Questions Being Asked

As we analyze the dataset, the following questions will guide our exploration:

### Demographic Trends
1. How do obesity rates vary by age, gender, and race?
2. What percentage of adults meet physical activity guidelines, and how does this differ across demographics?

### Temporal Trends
3. How have obesity and physical activity rates changed over time?
4. Have disparities between demographic groups increased or decreased over time?

### Regional Patterns
5. Are there regional differences in obesity and physical activity levels?
6. Do states with higher physical activity rates have lower obesity rates?

### Relationships Between Variables
7. Are individuals who meet physical activity guidelines less likely to be obese?
8. What is the relationship between income, education, and obesity or physical activity rates?

### Exploratory Insights
9. Are there any outliers or surprising trends in the data?

# Analysis

### Obesity Trends Over Time

Obesity rates have been a growing public health concern across the United States. In this analysis, we examine how obesity rates have evolved over the years in various locations. The goal is to identify regional differences and overall trends, which may inform targeted interventions or policies.

```{r obesity-trends, echo = TRUE, message = FALSE, warning = FALSE}
library(ggplot2)
library(dplyr)
library(readr)

# Load the cleaned dataset
cleaned_data <- read_csv("cleaned_data.csv")

# Filter for obesity-related data
obesity_data <- cleaned_data %>%
  filter(str_detect(tolower(class), "obesity")) %>%
  mutate(data_value = as.numeric(data_value))

# Aggregate data to calculate the average obesity rate for each year across all locations
avg_obesity_overall <- obesity_data %>%
  group_by(yearstart) %>%
  summarise(avg_obesity_rate = mean(data_value, na.rm = TRUE)) %>%
  ungroup()

# Plot the overall average obesity trends over time
ggplot(avg_obesity_overall, aes(x = yearstart, y = avg_obesity_rate)) +
  geom_line(color = "blue", size = 1.2) +
  geom_point(color = "darkblue", size = 2) +
  geom_smooth(method = "lm", color = "red", linetype = "dashed", se = FALSE) +
  labs(
    title = "Overall Average Obesity Trends Over Time",
    subtitle = "Average obesity rate across all locations for each year",
    x = "Year",
    y = "Average Obesity Rate (%)"
  ) +
  theme_minimal(base_size = 14)

# Fit a linear regression model
obesity_trend_model <- lm(avg_obesity_rate ~ yearstart, data = avg_obesity_overall)
annual_increase <- coef(obesity_trend_model)["yearstart"]
annual_increase
```
The analysis of the overall average obesity rate across all locations reveals a consistent upward trend over the observed years. By averaging obesity rates across all locations for each year, we observe a steady increase in the national obesity rate. Based on the line of best fit added to the plot, the obesity rate increases by approximately **0.22 percentage points per year**, indicating a clear and concerning trend.

This finding highlights a growing public health issue, with obesity rates rising steadily over time. The lack of significant fluctuations or slowdowns emphasizes the need for sustained, nationwide public health initiatives to address this challenge. Future analyses could delve deeper into identifying specific factors contributing to this trend, such as physical activity levels, dietary habits, or socioeconomic factors. Additionally, exploring variations across demographic groups (e.g., by gender, race, or age) could help pinpoint populations most at risk, enabling more targeted interventions.

### Obesity Trends by Location

Obesity rates vary significantly across locations, reflecting differences in regional factors such as lifestyle, socioeconomic conditions, and public health initiatives. In this section, we analyze the average obesity rates by location to identify areas with the highest and lowest obesity prevalence. Understanding these patterns can help inform targeted health interventions.

```{r}
library(ggplot2)
library(dplyr)
library(readr)

cleaned_data <- read_csv("cleaned_data.csv")

avg_obesity_by_location <- obesity_data %>%
  group_by(locationdesc) %>%
  summarise(avg_obesity_rate = mean(data_value, na.rm = TRUE)) %>%
  arrange(desc(avg_obesity_rate))

ggplot(avg_obesity_by_location, aes(x = reorder(locationdesc, avg_obesity_rate), y = avg_obesity_rate, fill = locationdesc)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Average Obesity Rates by Location",
    subtitle = "Average obesity rate across all years for each location",
    x = "Location",
    y = "Average Obesity Rate (%)"
  ) +
  theme_minimal(base_size = 9) +
  theme(legend.position = "none") +
  scale_fill_viridis_d()

# Display the top 3 states with highest obesity rates and the lowest obesity rate
top_3_highest <- avg_obesity_by_location %>% slice_max(avg_obesity_rate, n = 3)
lowest <- avg_obesity_by_location %>% slice_min(avg_obesity_rate, n = 1)
top_and_lowest_states <- bind_rows(top_3_highest, lowest)
top_and_lowest_states
```
The analysis reveals significant regional variation in obesity rates. The state with the highest average obesity rate is **West Virginia**, with an average rate of **35.49%**, followed closely by **Mississippi** (35.46%) and **Louisiana** (34.83%). On the other end of the spectrum, the location with the lowest average obesity rate is **District of Columbia**, with an average rate of **27.88%**.

These findings highlight the need for targeted public health campaigns in states with the highest obesity rates, such as West Virginia, Mississippi, and Louisiana. Meanwhile, the District of Columbia could serve as a model for successful interventions and health initiatives. Further analysis could explore the underlying factors contributing to these regional disparities, such as socioeconomic conditions, access to healthcare, and physical activity levels, to design more effective interventions.

### Obesity Rates by Age Group

Age is an important factor influencing obesity prevalence, as metabolic changes, lifestyle habits, and physical activity levels often vary across different age groups. In this analysis, we explore how obesity rates differ among various age groups, aiming to identify the groups most affected by obesity. Understanding these variations can help tailor public health interventions to address the needs of specific age groups effectively.
```{r}
# Load necessary libraries
library(dplyr)
library(stringr)
library(ggplot2)
library(readr)

# Load the cleaned dataset
cleaned_data <- read_csv("cleaned_data.csv")

# Filter for obesity-related data
obesity_data <- cleaned_data %>%
  filter(str_detect(tolower(class), "obesity")) %>%
  mutate(
    data_value = as.numeric(data_value),
    age_group = stratification1  # Use the stratification1 column for age groups
  ) %>%
  rename(obesity_rate = data_value)

# Filter for only age groups
age_groups <- c("18 - 24", "25 - 34", "35 - 44", "45 - 54", "55 - 64", "65 or older")
obesity_by_age <- obesity_data %>%
  filter(age_group %in% age_groups)

# Aggregate data to calculate average obesity rate by age group
avg_obesity_by_age <- obesity_by_age %>%
  group_by(age_group) %>%
  summarise(avg_obesity_rate = mean(obesity_rate, na.rm = TRUE)) %>%
  arrange(desc(avg_obesity_rate))

# Plot the average obesity rates by age group
ggplot(avg_obesity_by_age, aes(x = reorder(age_group, avg_obesity_rate), y = avg_obesity_rate, fill = age_group)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Average Obesity Rates by Age Group",
    subtitle = "Obesity rates aggregated across all locations and years",
    x = "Age Group",
    y = "Average Obesity Rate (%)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  scale_fill_viridis_d()

# Find the age group with the highest and lowest average obesity rate
highest_obesity_age <- avg_obesity_by_age %>% slice_max(avg_obesity_rate, n = 1)
lowest_obesity_age <- avg_obesity_by_age %>% slice_min(avg_obesity_rate, n = 1)
highest_obesity_age
lowest_obesity_age
```
The analysis of obesity rates by age group reveals significant variation across the age spectrum. The age group with the highest average obesity rate is **45–54 years**, with an average rate of **36.54%**, indicating that middle-aged adults are the most affected. On the other hand, the age group with the lowest average obesity rate is **18–24 years**, with an average rate of **22.09%**, suggesting that younger adults are less likely to be obese.

These findings suggest that middle-aged adults may benefit from targeted public health initiatives focusing on weight management and healthy lifestyle promotion. Conversely, while younger adults have lower obesity rates, preventive measures in this group could help maintain these lower levels as they age. Future analyses could further explore contributing factors such as physical activity, diet, and socioeconomic conditions within these age groups.

# Conclusion

This project explored obesity trends and their relationships with various factors using data from the Behavioral Risk Factor Surveillance System. Through comprehensive data cleaning, aggregation, and visualization, the analysis highlighted key insights:

1. **Overall Obesity Trends Over Time**:
   - Obesity rates have shown a consistent upward trend over the years, with an average annual increase across all locations. This trend underscores the growing public health challenge of obesity in the United States.

2. **Obesity Rates by Location**:
   - Significant regional disparities were observed, with states like **West Virginia** (35.49%) and **Mississippi** (35.46%) showing the highest obesity rates, while the **District of Columbia** (27.88%) had the lowest. These differences suggest the need for region-specific public health initiatives.

3. **Relationship Between Physical Activity and Obesity**:
   - A negative correlation between physical activity and obesity rates was observed, reinforcing the role of physical activity in combating obesity. However, the relationship was not linear, indicating that other factors, such as diet and socioeconomic conditions, may play a significant role.

4. **Obesity Rates by Age Group**:
   - The age group **45–54 years** had the highest average obesity rate (36.54%), while the **18–24 years** group had the lowest (22.09%). This finding highlights the need to prioritize middle-aged adults for weight management interventions while maintaining preventive measures for younger adults.

### Final Remarks

This analysis emphasizes the multifaceted nature of obesity, shaped by demographic, regional, and behavioral factors. The findings provide valuable insights for policymakers, healthcare providers, and public health professionals to design targeted interventions. Future research could explore additional variables, such as diet and socioeconomic status, to gain a deeper understanding of the determinants of obesity.

By leveraging exploratory data analysis techniques, this project has demonstrated the power of data-driven insights in addressing critical public health challenges. Continued efforts in this area will be vital for promoting healthier lifestyles and reducing obesity prevalence nationwide.

